---
title: "STA4020A - Assignment 1"
author: "Chesia Anyika"
date: "2024-05-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question

Consider the following two sets of data:

1)  Write the hypothesis to be tested

2)  Calculate the correction factor, Total SS, Treatment SS and Error SS

3)  Generate the ANOVA table

4)  At 5% and 10% levels of significance, test the hypothesis and make conclusions

# Case 1: Equal Replications

|                 | Treatment A | Treatment B | Treatment C |
|-----------------|-------------|-------------|-------------|
| **Replicate 1** | 23          | 42          | 47          |
| **Replicate 2** | 36          | 26          | 43          |
| **Replicate 3** | 31          | 47          | 43          |
| **Replicate 4** | 33          | 34          | 39          |

## 1. Write the Hypothesis to be Tested

The **Null hypothesis** is that all the means are equal, which is represented mathematically as: $H_0 : \mu_1 = \mu_2 = \mu_3$

The **Alternative hypothesis** is that at east one of the means is different from the other means, which is represented mathematically as either of the three:

$$
H_0 : \mu_1 = \mu_2 \neq \mu_3\\
H_0 : \mu_1 \neq \mu_2 = \mu_3\\
H_0 : \mu_1 \neq \mu_2 \neq \mu_3
$$

## 2. Calculate the correction factor, Total SS, Treatment SS and Error SS

### **Step 1: Input the table in r**

```{r}
#create vectors
Replicate <- c('Replicate 1', 'Replicate 2', 'Replicate 3', 'Replicate 4')
Treatment_A <- c(23, 36, 31, 33)
Treatment_B <- c(42, 26, 47, 34)
Treatment_C <- c(47, 43, 43, 39)

#create table
case1 <- data.frame(Replicate, Treatment_A, Treatment_B, Treatment_C)

#View
case1
```

I then computed the row-wise sums as $Y_i$, and added these to the table.

```{r}
#compute rowsums
Sums <- c('Y_i', sum(case1$Treatment_A), sum(case1$Treatment_B), sum(case1$Treatment_C))

#rbind to table
case1 <- rbind(case1, Sums)

#View
case1
```

### Step 2: Correction factor

The correction factor is computed as per the formula:

$$
CF = \frac{Y..^2}{rt}
$$

> Where:
>
> $Y..$ is the sum of the row sums.
>
> $r$ is the number of replicates.
>
> $t$ is the number of treatments.

I first defined my parameters $r$ and $t$, and computed $Y..$ , the sum of the row sums, as follows:

```{r}
#define parameters
r <- 4
t <- 3

#compute the sum
Ydot <- 123 + 149 + 172

#View Result
cat('Y.. = ', Ydot)
```

I then computed the formula:

```{r}
#Formula
CF <- ((Ydot)^2/(r*t))

#View
CF
```

The **Correction Factor** is $16,428$

### Step 3: Total SS

Total SS is computed as per the formula:

$$
TotalSS = \sum Y^2_{ij} - CF
$$

To compute $\sum Y^2_{ij}$, I computed the individual squares of each element and summed them up

```{r}
#compute individual squares
TA.sq <- Treatment_A^2
TB.sq <- Treatment_B^2
TC.sq <- Treatment_C^2

#sum all the squares
Y2sum <- sum(TA.sq, TB.sq, TC.sq)

#View the result
Y2sum
```

I then computed the TotalSS as per the formula:

```{r}
#compute TotalSS
TotalSS <- Y2sum - CF

#View results
TotalSS
```

The **TotalSS** is $680$.

### Step 4: Treatment SS

The **Treatment SS (TRT SS)** is computed as per the formula:

$$
 TRTSS = \sum \frac{Y^2_i}{r} - CF
$$

I computed this as per the formula, as follows

```{r}
#compute the TRTSS
Yi <- c(123, 149, 172)
TRTSS <- sum((Yi^2)/r)-CF

#View the result
TRTSS
```

The **Treatment SS** **(TRTSS)** is $300.5$.

### Step 5: Error SS

The **Error SS** is computed as per the formula:

$$
ErrorSS = TotalSS - TRTSS
$$

I computed this as follows:

```{r}
#compute the formula
ErrorSS <- TotalSS - TRTSS

#View the result
ErrorSS
```

The **ErrorSS** is $379.5$.

## 3. Generate the ANOVA table

The Anova table requires **Degrees of Freedom**, the **Sum of Squares(Total, Treatment, Error)**, the **Mean Squares** and the **F-Statistic**.

First I computed the degrees of freedom, as per the formulas:

$$
DF_{trt} = t - 1\\
DF_{err} = t \times (r - 1)\\
DF_{total} = \frac{DF_{trt}}{DF_{err}} 
$$

> Where:
>
> $DF_{trt}$ is the **Treatment Degrees of Freedom**
>
> $DF_{err}$ is the **Error Degrees of Freedom**
>
> $DF_{total}$ is the **Total Degrees of Freedom**.

```{r}
# Compute degrees of freedom
Df_trt <- t - 1
Df_err <- t * (r - 1)
Df_tot <- Df_trt + Df_err

#View results
cat('\n Treatment Df: ', Df_trt, '\n Error Df: ', Df_err, '\n Total Df: ', Df_tot)
```

I then computed the Mean Squares (MS), as per the formulas:

$$
MS_{trt} = \frac{SS_{trt}}{DF_{trt}}\\
MS_{err} = \frac{SS_{err}}{DF_{err}}
$$

> Where:
>
> $MS_{trt}$ is the **Treatment Mean Squares.**
>
> $MS_{err}$ is the **Error Mean Squares**.

This is computed as follows:

```{r}
# Compute Mean Square (MS)
MS_trt <- TRTSS / Df_trt
MS_err <- ErrorSS / Df_err

#View results
cat('\n Treatment MS: ', MS_trt, '\n Error MS: ', MS_err)
```

I then computed the F-statistic as per the formula:

$$
F_{value} = \frac{MS_{trt}}{MS_{err}}
$$

This is computed as follows:

```{r}
# Compute F-value for Treatment
F_value <- MS_trt / MS_err

#View results
cat('F value: ', F_value)
```

I then input all the values into a data-frame, to create the anova table as follows:

```{r}
# Create ANOVA table
anova_table <- data.frame(
  Sources_of_Variation = c("Treatment", "Error", "Total"),
  Df = c(Df_trt, Df_err, Df_tot),
  SS = c(TRTSS, ErrorSS, TotalSS),
  MS = c(MS_trt, MS_err, NA),  # MS for Total is not applicable
  F = c(F_value, NA, NA)  # F for Error and Total is not applicable
)

# Print ANOVA table
print(anova_table, row.names = FALSE)

```

## 4. At 5% and 10% levels of significance, test the hypothesis and make conclusions

First, I defined my degrees of freedom.

```{r}
# Degrees of freedom for critical F-value
Df1 <- Df_trt
Df2 <- Df_err
```

### Step 1: At 5% Significance

I used the `qf()` function to obtain the Critical F value at a 5% level of significance. I then compared the obtained F-statistic to the F_value, to make conclusions on the null hypothesis.

```{r}
# Critical F-value at 5% significance level
critical_F_5 <- qf(0.95, Df1, Df2)

# Test hypothesis at 5% significance level
if (F_value > critical_F_5) {
  cat("\n F_value: ", F_value, "\n Critical Value: ", critical_F_5, "\n ", "\n Null hupothesis rejected")
} else {
  cat("\n F_value: ", F_value, "\n Critical Value: ", critical_F_5, "\n ", "\n Fail to reject null hypothesis")
}
```

The F-value (3.56) is **less than** the critical value (4.26), thus at a 95% confidence level we fail to reject the null hypothesis stated as follows: $H_0: \mu_1 = \mu_2 = \mu_3$. Thus, there are no significant differences among the means.

### Step 2: At 10% Significance

I repeated the procedure at 10% significance level.

```{r}
# Critical F-value at 10% significance level
critical_F_10 <- qf(0.90, Df1, Df2)

# Test hypothesis at 10% significance level
if (F_value > critical_F_10) {
  cat("\n F_value: ", F_value, "\n Critical Value: ", critical_F_10, "\n ", "\n Null hupothesis rejected")
} else {
  cat("\n F_value: ", F_value, "\n Critical Value: ", critical_F_5, "\n ", "\n Fail to reject null hypothesis")
}
```

The F_value (3.56) is **greater than** the Critical value (3.01), thus at a 90% confidence level we reject the null hypothesis stated as $H_0: \mu_1 = \mu_2 = \mu_3$. Thus, there is at least one statistically significant difference among the means.

# Case 2: Unequal Replications

|                 | Treatment A | Treatment B | Treatment C | Treatment D |
|-----------------|-------------|-------------|-------------|-------------|
| **Replicate 1** | 2.0         | 1.7         | 2.0         | 2.1         |
| **Replicate 2** | 2.2         | 1.9         | 2.4         | 2.2         |
| **Replicate 3** | 1.8         | 1.5         | 2.7         | 2.2         |
| **Replicate 4** | 2.3         |             | 2.5         | 1.9         |
| **Replicate 5** | 1.7         |             | 2.4         |             |

## 1. Write the Hypothesis to be Tested

The **Null hypothesis** is that all the means are equal, which is represented mathematically as: $H_0 : \mu_1 = \mu_2 = \mu_3 = \mu_4$

The **Alternative hypothesis** is that at east one of the means is different from the other means, which is represented mathematically as follows:

$$
H_0 : \mu_1 \neq \mu_2 \neq \mu_3 \neq \mu_4
$$

## 2. Calculate the correction factor, Total SS, Treatment SS and Error SS

### Step 1: Input the Table

First I created the above table as follows:

```{r}
#Input the values
replicates <- c('Replicate 1', 'Replicate 2', 'Replicate 3', 'Replicate 4', 'Replicate 5')
treatment_A <- c(2.0, 2.2, 1.8, 2.3, 1.7)
treatment_B <- c(1.7, 1.9, 1.5, NA, NA)
treatment_C <- c(2.0, 2.4, 2.7, 2.5, 2.4)
treatment_D <- c(2.1, 2.2, 2.2, 1.9, NA)

#create table
case2 <- data.frame(replicates, treatment_A,treatment_B, treatment_C, treatment_D)

#View
case2
```

I then computed $Y_i$, the row sums, as follows.

```{r}
#compute rowsums
sums <- c('Y_i', sum(case2$treatment_A), sum(1.7, 1.9, 1.5), sum(case2$treatment_C), sum(2.1, 2.2, 2.2, 1.9))

#rbind to table
case2 <- rbind(case2, sums)

#View
case2
```

### Step 2: Correction Factor

The correction factor is computed as per the formula:

$$
 CF = \frac{Y..^2}{\sum r_i}
$$

> Where:
>
> $Y..$ is the sum of the row sums.
>
> $\sum r_i$ is the number of replicates. (there ar 17 in this case)

I first defined my parameters $\sum r_i$, and computed $Y..$ , the sum of the row sums, as follows:

```{r}
#define parameters
ri.sum <- 17

#compute the sum
Ydot2 <- 10 + 5.1 + 12 + 8.4

#View Result
cat('Y.. = ', Ydot2)
```

I then computed the formula

```{r}
#Formula
CF <- ((Ydot2)^2/(ri.sum))

#View
CF
```

The **Correction Factor** is $74.13$

### Step 3: Total SS

Total SS is computed as per the formula:

$$
TotalSS = \sum Y^2_{ij} - CF
$$

To compute $\sum Y^2_{ij}$, I computed the individual squares of each element and summed them up.

```{r}
Y2.sum <- sum(2.0^2, 2.2^2, 1.8^2, 2.3^2, 1.7^2, 1.7^2, 1.9^2, 1.5^2, 2.0^2, 2.4^2, 2.7^2, 2.5^2, 2.4^2, 2.1^2, 2.2^2, 2.2^2, 1.9^2)
```

I then computed the formula as follows:

```{r}
#compute formula
TotalSS <- Y2.sum - CF

#View result
TotalSS
```

The **Total Sum of Squares** is $1.638$

### Step 4: Treatment SS

The Treatment SS (TRT SS) is computed as per the formula:

$$
TRTSS = \sum \frac{Y^2_i}{r_i} - CF
$$

First, I defined my parameters, then computed the formula as follows:

```{r}
#define parameters
ri <- c(5, 3, 5, 4)
Y_i <- c(10, 5.1, 12, 8.4)

#compute the formula
TRTSS <- sum(((Y_i)^2)/ri) - CF

#View result
TRTSS
```

The **Treatment Sum of Squares** is $0.98$.

### Step 5: Error SS

The Error SS is computed as per the formula:

$$
ErrorSS = TotalSS - TRTSS
$$

I computed this as follows:

```{r}
#compute the formula
ErrorSS <- TotalSS - TRTSS

#view
ErrorSS
```

The **Error Sum of Squares** is $0.66$.

## 3. Generate the ANOVA table

The Anova table requires **Degrees of Freedom, the Sum of Squares(Total, Treatment, Error), the Mean Squares** and the **F-Statistic.**

First I computed the degrees of freedom, as per the formulas:

$$
DF_{trt} = t - 1\\
DF_{err} = \sum r_i - t\\
DF_{total} = \sum r_i - 1 
$$

> Where:
>
> $DF_{trt}$ is the Treatment Degrees of Freedom
>
> $DF_{err}$ is the Error Degrees of Freedom
>
> $DF_{total}$ is the Total Degrees of Freedom.

```{r}
#define parameters
t <- 4
ri <- 17

# Compute degrees of freedom
Df_trt <- t - 1
Df_err <- ri - t
Df_tot <- ri - 1

#View results
cat('\n Treatment Df: ', Df_trt, '\n Error Df: ', Df_err, '\n Total Df: ', Df_tot)
```

I then computed the **Mean Squares (MS)**, as per the formulas:

$$
MS_{trt} = \frac{SS_{trt}}{DF_{trt}}\\
MS_{err} = \frac{SS_{err}}{DF_{err}}
$$

> Where:
>
> $MS_{trt}$ is the Treatment Mean Squares.
>
> $MS_{err}$ is the Error Mean Squares.

This is computed as follows:

```{r}
# Compute Mean Square (MS)
MS_trt <- TRTSS / Df_trt
MS_err <- ErrorSS / Df_err

#View results
cat('\n Treatment MS: ', MS_trt, '\n Error MS: ', MS_err)
```

I then computed the F-statistic as per the formula:

$$
F_{value} = \frac{MS_{trt}}{MS_{err}}
$$

This is computed as follows:

```{r}
# Compute F-value for Treatment
F_value <- MS_trt / MS_err

#View results
cat('F value: ', F_value)
```

I then input this into a dataframe to create the anova table as follows:

```{r}
# Create ANOVA table
anova_table2 <- data.frame(
  Sources_of_Variation = c("Treatment", "Error", "Total"),
  Df = c(Df_trt, Df_err, Df_tot),
  SS = c(TRTSS, ErrorSS, TotalSS),
  MS = c(MS_trt, MS_err, NA),  # MS for Total is not applicable
  F = c(F_value, NA, NA)  # F for Error and Total is not applicable
 
)

# Print ANOVA table
print(anova_table2, row.names = FALSE)

```

## 4. At 5% and 10% levels of significance, test the hypothesis and make conclusions

First I defined my degrees of freedom.

```{r}
# Degrees of freedom for critical F-value
Df1 <- Df_trt
Df2 <- Df_err
```

### Step 1: At 5% Significance Level

I used the `qf()` function to obtain the Critical F value at a 5% level of significance. I then compared the obtained F-statistic to the F_value, to make conclusions on the null hypothesis.

```{r}
# Critical F-value at 5% significance level
critical_F_5 <- qf(0.95, Df1, Df2)

# Test hypothesis at 5% significance level
if (F_value > critical_F_5) {
  cat("\n F_value: ", F_value, "\n Critical Value: ", critical_F_5, "\n ", "\n Null hupothesis rejected")
} else {
  cat("\n F_value: ", F_value, "\n Critical Value: ", critical_F_5, "\n ", "\n Fail to reject null hypothesis")
}
```

The F_value (6.41) is **greater than** the Critical value (3.41), thus at a 90% confidence level **we reject the null hypothesis** stated as $H_0: \mu_1 = \mu_2 = \mu_3$. Thus, there is at least one statistically significant difference among the means.

### Step 2: At 10% Significance Level

I repeated the procedure at 10% significance level.

```{r}
# Critical F-value at 10% significance level
critical_F_10 <- qf(0.90, Df1, Df2)

# Test hypothesis at 10% significance level
if (F_value > critical_F_10) {
  cat("\n F_value: ", F_value, "\n Critical Value: ", critical_F_10, "\n ", "\n Null hupothesis rejected")
} else {
  cat("\n F_value: ", F_value, "\n Critical Value: ", critical_F_5, "\n ", "\n Fail to reject null hypothesis")
}
```

The F_value (6.42) is **greater than** the Critical value (2.56), thus at a 90% confidence level **we reject the null hypothesis** stated as $H_0: \mu_1 = \mu_2 = \mu_3$. Thus, there is at least one statistically significant difference among the means.
